---
title: "Measles Global Analysis"
format:
  dashboard:
    orientation: columns
    embed-resources: true
    theme: journal
    scrolling: false
    logo: measles.png
    nav-buttons:
      - icon: github
        href: https://github.com/aryanajithnair/STAT-431
      - icon: envelope
        href: mailto:ajithnai@calpoly.edu
      - icon: envelope
        href: mailto:amdong@calpoly.edu
      - icon: envelope
        href: mailto:nacartwr@calpoly.edu
      - icon: envelope
        href: mailto:ztam@calpoly.edu
        
server: shiny
---

# Global Measles Cases
```{css}
.bslib-value-box .value-box-area {
  padding: 0.5rem;
}

.bslib-value-box .bi {
  font-size: 1.5rem !important;
}
```

```{r}
#| label: packages-setup-and-load-data
#| message: false
#| warning: false
library(httr)
library(tidyjson)
library(jsonlite)
library(tidyverse)
library(stringr)
library(gt)
library(paletteer)
library(ggridges)
library(broom)
library(gganimate)
library(leaflet)
library(shiny)

cases_month <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-06-24/cases_month.csv')
cases_year <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-06-24/cases_year.csv')
meta_df <- readr::read_csv("https://raw.githubusercontent.com/zeetahhh/measles-analysis/main/mg%20copy.csv")

trend_label <- function(current, previous) {
  pct_change <- (current - previous) / previous * 100
  case_when(
    pct_change < -20 ~ "declining",
    pct_change >  20 ~ "surging",
    .default = "stable"
  )
}
```

<h4>Measles Cases from 2012-2025</h4>

::: {.panel-tabset}
## Figure 1: Measles Cases Over Time {width="50%"}
```{r}
#| label: slider-ui
sliderInput(inputId = "year", label = "Select a year:", min = 2012, max = 2025, value = 2012, sep = "")

plotOutput(outputId = "yearPlot")
```

```{r}
#| context: server
library(tidyverse)
cases_month <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-06-24/cases_month.csv')
monthly_sum <- cases_month %>% group_by(year, month) %>% summarize(cases = sum(measles_total, na.rm = TRUE))

output$yearPlot <- renderPlot({
  plotdata <- monthly_sum %>% filter(year == input$year)
  ggplot(data = plotdata, aes(x = month, y = cases)) +
    geom_line() +
    scale_x_continuous(breaks = seq(1, 12, by = 1), labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")) +
    labs(title = paste("Total Measles Cases by Month (", input$year, ")", sep = "")) +
    xlab("Month") +
    ylab("Total Cases") +
    theme_bw()
  })
```

## Table 1: Measles by Year
```{r}
#| label: gt-measles-table-rq1
# --code-- Insert table 1 of global cases by year

yearly_summary <- cases_year |>
  group_by(year) |>
  summarise(
    total_cases = sum(measles_total, na.rm = TRUE),
    n_countries = n_distinct(country),
    .groups = "drop"
  ) |>
  arrange(year) |>
  mutate(
    prev_total = lag(total_cases),
    pct_change = round((total_cases - prev_total) / prev_total * 100, 1),
    trend = trend_label(total_cases, prev_total)
  ) |>
  select(year, total_cases, n_countries, pct_change, trend)

yearly_summary |>
  select(year, total_cases, n_countries, pct_change, trend) |>
  gt() |>
  cols_label(
    year = md("**Year**"),
    total_cases = md("**Total Cases**"),
    n_countries = md("**Countries Reporting**"),
    pct_change = md("**% Change from Prior Year**"),
    trend = md("**Trend**")
  ) |>
  fmt_number(columns = total_cases, use_seps = TRUE, decimals = 0) |>
  fmt_number(columns = n_countries, use_seps = TRUE, decimals = 0) |>
  fmt_number(columns = pct_change, decimals = 1, suffix = "%") |>
  data_color(
    columns = pct_change,
    fn = scales::col_numeric(
      palette = c("#D9565C", "#FFFFFF", "#647D4B"),
      domain  = c(-100, 200)
    )
  ) |>
  tab_style(
    style = cell_text(color = "#D9565C", weight = "bold"),
    locations = cells_body(columns = trend, rows = trend == "declining")
  ) |>
  tab_style(
    style = cell_text(color = "#647D4B", weight = "bold"),
    locations = cells_body(columns = trend, rows = trend == "surging")
  ) |>
  tab_style(
    style = cell_text(color = "#848C80"),
    locations = cells_body(columns = trend, rows = trend == "stable")
  ) |>
  tab_spanner(
    label = md("**Case Counts**"),
    columns = c(total_cases, n_countries)
  ) |>
  tab_spanner(
    label = md("**Year-over-Year Change**"),
    columns = c(pct_change, trend)
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_spanners()
  ) |>
  tab_options(
    table.width = pct(100),
    column_labels.font.weight = "bold",
    heading.align = "left",
    table.border.top.color = "black",
    row.striping.include_table_body = TRUE
  ) |>
  sub_missing(missing_text = "—")
```

:::

## Column - rightside (cards, measles detection) {width="50%"}

### Row - RQ1 Value Boxes {height="165px"}
```{r}
yearly_totals  <- cases_year |>
   group_by(year) |>
   summarise(total = sum(measles_total, na.rm = TRUE))

 highest_year   <- yearly_totals |> slice_max(total, n = 1)
 lowest_year    <- yearly_totals |> slice_min(total, n = 1)
 max_pct_change <- yearly_totals |>
   mutate(pct_change = (total - lag(total)) / lag(total) * 100) |>
   slice_max(abs(pct_change), n = 1) |>
   pull(pct_change) %>% round(digits= 2)
```

::: {.valuebox icon="arrow-up-circle" color="danger"}
**Highest single-year cases**

`{r} highest_year$year`

`{r} str_c(format(highest_year$total, scientific = FALSE), "cases", sep = " ")`
:::

::: {.valuebox icon="arrow-down-circle" color="success"}
**Lowest single-year cases**

`{r} lowest_year$year`

`{r} str_c(format(lowest_year$total, scientific = FALSE), "cases", sep = " ")`
:::

::: {.valuebox icon="percent" color="info"}
**Largest year-over-year % change**

`{r} str_c(max_pct_change, "%")`
:::


### Row - Table 2 

::: {.card title="Table 2: Detection Method Proportions" fill="false"}
```{r}
#| label: detection-ui

library(shiny)
library(gt)

checkboxGroupInput(
  inputId = "detection_types",
  label = "Select Detection Methods:",
  choices = c("Clinical" = "clin_prop",
              "Epi-Linked" = "epi_prop",
              "Lab Confirmed" = "lab_prop"),
  selected = c("clin_prop", "epi_prop", "lab_prop")
)
gt_output("detection_table")
```

```{r}
#| label: detection-server
#| context: server

library(tidyverse)
library(gt)
cases_year <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-06-24/cases_year.csv')
meta_df <- readr::read_csv("https://raw.githubusercontent.com/zeetahhh/measles-analysis/main/mg%20copy.csv")
cases_filtered <- cases_year %>% filter(!is.na(measles_clinical),
                                     !is.na(measles_total),
                                     !is.na(measles_epi_linked),
                                     !is.na(measles_lab_confirmed))

yearly_totals_det <- cases_filtered %>% group_by(year) %>%
  summarise(
    total = sum(measles_total),
    clinical = sum(measles_clinical),
    epi = sum(measles_epi_linked),
    lab = sum(measles_lab_confirmed),
    .groups = "drop") %>%
  mutate(
    clin_prop = clinical / total,
    epi_prop = epi / total,
    lab_prop = lab / total
  ) %>%
  select(year, clin_prop, epi_prop, lab_prop)


output$detection_table <- render_gt({
  req(input$detection_types)
  selected_cols <- input$detection_types
  table_data <- yearly_totals_det |> select(year, all_of(selected_cols))

  col_labels <- list(year = md("**Year**"))
  if ("clin_prop" %in% selected_cols) col_labels$clin_prop <- md("**Clinical**")
  if ("epi_prop"  %in% selected_cols) col_labels$epi_prop  <- md("**Epi-Linked**")
  if ("lab_prop"  %in% selected_cols) col_labels$lab_prop  <- md("**Lab Confirmed**")

  tbl <- table_data |>
    gt() |>
    cols_label(.list = col_labels) |>
    fmt_percent(columns = all_of(selected_cols), decimals = 1) |>
    tab_options(table.width = pct(100), column_labels.font.weight = "bold",
                heading.align = "left", table.border.top.color = "black",
                row.striping.include_table_body = TRUE)

  if (length(selected_cols) > 1) {
    tbl <- tbl |>
      tab_spanner(label = md("**Proportion of Total Cases by Detection Method**"),
                  columns = all_of(selected_cols)) |>
      tab_style(style = cell_text(weight = "bold"), locations = cells_column_spanners())
  }

  for (col in selected_cols) {
    tbl <- tbl |>
      tab_style(style = cell_fill(color = "#D4EFDF"),
                locations = cells_body(columns = all_of(col),
                  rows = table_data[[col]] == do.call(pmax, c(table_data[selected_cols], na.rm = TRUE))))
  }
  tbl
})
```

:::

# Madagascar

## Column - Country Context {width="30%"} 

::: {.card title="Madagascar's Outbreak"}
Madagascar is the most striking outlier in the global measles dataset, its 2018–2019 outbreak represents the largest single-country spike in the entire dataset. In 2012–2017, fewer than 100 cases were recorded per year, but cases surged to over 213,000 in 2018–2019 (a <mark>~1,700% increase</mark>). Following this peak, cases declined dramatically back to fewer than 100.
A primary driver of this outbreak was chronically low vaccination coverage. Madagascar's measles immunization rates had fallen well below the 95% threshold needed for herd immunity in the years prior, leaving a large susceptible population, particularly among young children in rural areas with limited healthcare access. When the virus was reintroduced, it spread rapidly through these unvaccinated communities. The outbreak ultimately prompted a nationwide emergency vaccination campaign, which is widely credited with the sharp decline seen in subsequent years.
:::

## Column - Madagascar Visualizations {width="70%"}

### Row - Animated Map {height="55%"}
::: {.card title="Figure 2: Madagascar Measles Cases Over Time"}
```{r}
#| label: mad-plotly
#| warning: false
#| message: false
madagascar <- cases_year %>% filter(country == "Madagascar")

ggplot(data = madagascar, aes(x = year, y = total_population / 1000000, group = 1)) +
  geom_line() +
  geom_point() +
  scale_y_continuous(breaks = seq(20, 35, by = 5)) +
  scale_x_continuous(breaks = seq(2012, 2025, by = 2)) +
  labs(title = "Madagascar's Population by Year") +
  xlab("Year (2012 - 2025)") +
  ylab("Population (in Millions)") +
  transition_reveal(year) +
  theme_bw(base_size= 14)
```
:::


### Row - Interactive Maps {height="45%"}

::: {.panel-tabset}

## All Cities {height="50%"}

::: {.card title="Figure 3: Madagascar City Populations"}
```{r}
#| warning: false
#| message: false
madagascarLat <- -18.77
madagascarLon <- 46.85
madagascar_cities <- meta_df

pal <- colorNumeric(palette = "plasma", domain = madagascar_cities$population)

leaflet(height = 350) %>%
  setView(lat = madagascarLat, lng = madagascarLon, zoom = 6) %>%
  addProviderTiles("Esri.WorldStreetMap") %>%
  addCircleMarkers(
    data = madagascar_cities,
    lat = ~lat,
    lng = ~lng,
    radius = 6,
    fillColor = ~pal(population),
    fillOpacity = 0.8,
    color = "white",
    weight = 1,
    popup = ~paste0("<b>", city, "</b><br>Population: ", scales::comma(population))
  ) %>%
  addLegend(
    pal = pal,
    values = madagascar_cities$population,
    title = "Population",
    position = "bottomright"
  )
```
:::


## Excluding Capital

::: {.card title="Figure 4: Madagascar City Populations (Excluding Antananarivo)"}
```{r}
#| warning: false
#| message: false
madagascarWithoutAnatananarivo <- madagascar_cities |>
  filter(city != "Antananarivo")

pal2 <- colorNumeric(palette = "plasma", domain = madagascarWithoutAnatananarivo$population)

leaflet(height = 350) %>%
  setView(lat = madagascarLat, lng = madagascarLon, zoom = 6) %>%
  addProviderTiles("Esri.WorldStreetMap") %>%
  addCircleMarkers(
    data = madagascarWithoutAnatananarivo,
    lat = ~lat,
    lng = ~lng,
    radius = 6,
    fillColor = ~pal2(population),
    fillOpacity = 0.8,
    color = "white",
    weight = 1,
    popup = ~paste0("<b>", city, "</b><br>Population: ", scales::comma(population))
  ) %>%
  addLegend(
    pal = pal2,
    values = madagascarWithoutAnatananarivo$population,
    title = "Population",
    position = "bottomright"
  )
```
:::

:::