---
title: "checkpoint 4"
author: "Zita Tam, Natalie Carthwright, Aryan Nair, Amanda Dong"
format:
  html:
    embed-resources: true
    code-fold: true
    theme: flatly
---

# load in packages and datasets
```{r}
#| label: packages-setup
#| message: false
#| warning: false
library(tidyverse)
library(paletteer)
library(gt)
library(stringr)
cases_month <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-06-24/cases_month.csv')
cases_year <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-06-24/cases_year.csv')
meta_df <- read_csv("https://raw.githubusercontent.com/amandaadong/project1/main/metadf.csv")
```

# Research question: Has the total number of reported measles cases increased or decreased over time globally? 

## Custom Function
```{r}
trend_label <- function(current, previous) {
  pct_change <- (current - previous) / previous * 100
  case_when(
    pct_change < -20 ~ "declining",
    pct_change >  20 ~ "surging",
    .default = "stable"
  )
}

yearly_summary <- cases_year |>
  group_by(year) |>
  summarise(
    total_cases = sum(measles_total, na.rm = TRUE),
    n_countries = n_distinct(country),
    .groups = "drop"
  ) |>
  arrange(year) |>
  mutate(
    prev_total = lag(total_cases),
    pct_change = round((total_cases - prev_total) / prev_total * 100, 1),
    trend = trend_label(total_cases, prev_total)
  ) |>
  select(year, total_cases, n_countries, pct_change, trend)
```

## Visualization using the Custom Function (complex geometry)

```{r}
#| warning: false
#| fig-alt: "Lollipop chart showing the percent change in total measles cases across the world for each year (2012 - 2025). The largest percent increase was in 2022, while the largest decrease was in 2020."

ggplot(yearly_summary, aes(x = year, xend = year, y = 0, yend = pct_change,
                           color = trend)) +
  geom_segment(size = 1.3, alpha = 0.9) +
  geom_point(aes(y = pct_change)) +
  theme_minimal() +
  theme(panel.border = element_blank()) +
  scale_x_continuous(breaks = yearly_summary$year) +
  scale_y_continuous(breaks = seq(-100, 200, by = 50)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
    labs(title = "Global Measles Cases Over Time", subtitle = "Percent Change Year-to-Year (2012-2025)") +
  xlab("Year") + ylab("Percent Change (%)") +
  scale_color_manual(values = c(surging = "#647D4BFF",
                      declining = "#D9565CFF",
                      stable = "#848C80"))
```

# Research Question: Does the method of finding measles cases differ according to the year?

## Data Summary

```{r}
propsbyyear <- cases_year %>% filter(!is.na(measles_clinical),
                                     !is.na(measles_total),
                                     !is.na(measles_epi_linked),
                                     !is.na(measles_lab_confirmed)) %>% 
  group_by(year) %>%
  summarise(clin_prop = (sum(measles_clinical) / sum(measles_total)),
         epi_prop = (sum(measles_epi_linked) / sum(measles_total)),
         lab_prop = (sum(measles_lab_confirmed) / sum(measles_total)))

plot_long <- propsbyyear %>%
  pivot_longer(cols = c(clin_prop, epi_prop, lab_prop),
               names_to = "method",
               values_to = "proportion")
```

## Visualization

```{r}
#| fig-alt: "A stacked bar graph showing the proportions of total measles cases that were either clinically-compatible cases, epidemiologically-linked cases, or laboratory-confirmed cases. Each bar represents one year. Lab-confirmed cases have the largest proportion across years, while epidemiologically-linked cases are generally the snallest proportion."
ggplot(plot_long, aes(x = year, y = proportion, fill = method)) +
  geom_bar(position = "fill", stat = "identity") +
  scale_x_continuous(breaks = plot_long$year) +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 8)) +
  labs(title = "Measles Detection Method by Year",
       subtitle = "Clinical vs. Epi-linked vs. Lab Cases") +
  xlab("Year") +
  ylab("Proportion") +
  scale_fill_paletteer_d("nationalparkcolors::Acadia", 
                         labels = c(clin_prop = "Clinical",
                                 epi_prop = "Epi-Linked",
                                 lab_prop = "Lab Confirmed"))
```

# Research question: Is there an association between a country's air quality and number of measles cases?

## Check meta dataset
```{r}
meta_df |>
  select(iso3, country, year, gdp_per_capita_nominal, 
         measles_incidence_rate_per_1000000_total_population,
         measles_lab_confirmed, measles_total) |>
  glimpse()

# Check for NAs
meta_df |>
  summarize(
    n_total = n(),
    n_missing_gdp = sum(is.na(gdp_per_capita_nominal)),
    n_missing_incidence = sum(is.na(measles_incidence_rate_per_1000000_total_population)),
    n_missing_lab = sum(is.na(measles_lab_confirmed)),
    n_missing_total = sum(is.na(measles_total)))
```

## Visualization
```{r}
#| message: false
#| warning: false
#| fig-alt: "A scatter plot with Air Quality Index for a country and the country's average annual measles cases count on a log scale. The linear regression appears to have a slight downward trend, which we will check by fitting a model"
meta_df %>%
  filter(!is.na(total_aqi), !is.na(total_suspected_measles_rubella_cases)) %>%
  group_by(country, total_aqi) %>%
  summarise(avg_cases = mean(total_suspected_measles_rubella_cases, na.rm = TRUE)) %>%
  ggplot(aes(x = total_aqi, y = avg_cases)) +
  geom_point(alpha = 0.6, color = "dodgerblue") +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  scale_y_log10(labels = scales::comma) +
  labs(
    title = "Air Quality Index vs. Average Measles Cases by Country",
    subtitle = "Each point represents one country; cases shown on log scale",
    x = "Air Quality Index (higher = worse air quality)",
    y = "Average Annual Measles Cases (log scale)",
    caption = "Sources: API Ninjas AQI data; TidyTuesday measles dataset"
  )
```

## Fitting the Model
```{r}
aqi_model <- meta_df |>
  filter(!is.na(total_aqi), !is.na(total_suspected_measles_rubella_cases)) |>
  group_by(country, total_aqi) |>
  summarise(avg_cases = mean(total_suspected_measles_rubella_cases, na.rm = TRUE),
            .groups = "drop") |>
  mutate(log_avg_cases = log10(avg_cases + 1))

model <- lm(log_avg_cases ~ total_aqi, data = aqi_model)

wide_model <- broom::tidy(model, conf.int = TRUE) %>%
  mutate(
    term = recode(term,
      "(Intercept)" = "Intercept",
      "total_aqi" = "Air Quality Index"
    ), conf_int = str_c("(", round(conf.low, 3), ", ", round(conf.high, 3), ")")
  ) %>%
  select(term, estimate, std.error, statistic, p.value, conf_int) %>%
  rename(c(Term = term, Estimate = estimate, SE = std.error, Statistic = statistic,
           "P-value" = p.value, "Conf. Interval" = conf_int))

wide_model |>
  gt() |>
  fmt_number(columns = -"Conf. Interval", decimals = 3) |>
  tab_style(style = cell_text(weight = "bold"),
                              locations = cells_column_labels()) |>
  tab_style(style = cell_text(weight = "bold"),
            locations = cells_body(
              columns = Term,
              rows = Term %in% c("Intercept", "Air Quality Index")
            )) |>
  tab_caption(caption = "Table 1. Linear Regression coefficients for the association between measles cases and AQI.")
```



```{r}
meta_df |>
  filter(!is.na(total_aqi), !is.na(measles_total)) |>
  group_by(country, total_aqi) |>
  summarise(avg_cases = mean(measles_total, na.rm = TRUE)) |>
  ggplot(aes(x = total_aqi, y = avg_cases + 1)) +
  geom_hex(bins = 15) +
  scale_y_log10(labels = scales::comma) +
  scale_fill_gradient(low = "#E6F1FA", high = "#0F6E56", name = "Countries") +
  labs(
    x = "Air Quality Index",
    y = "Avg annual measles cases (log scale)"
  ) +
  theme_minimal()
```

