---
title: "project1-checkpoint-3"
author: Natalie, Cartwright, Amanda Dong, Aryan Nair, Zita Tam
format: html
---

load in packages and dataset:
```{r}
#| label: packages-setup
#| message: false
#| warning: false
library(httr)
library(tidyjson)
library(jsonlite)
library(tidyverse)
library(stringr)
cases_month <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-06-24/cases_month.csv')
cases_year <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-06-24/cases_year.csv')
```

```{r}
#| echo: false
my_key <- "YOUR_API_KEY_HERE"
```

**Example with Canada 2026**

To start, I am going to use the API to access the GDP data for Canada in 2026.
```{r}
#| eval: false
base_url <- "https://api.api-ninjas.com/v1/gdp"
country <- "Canada"
year <- "2026"
url <- str_c(base_url, "?country=", country, "&year=", year)
canada_raw <- GET(url, add_headers("X-Api-Key" = my_key))
```

We now have the raw data, so I will change it into JSON format and parse in to a tidy dataframe.
```{r}
#| eval: false
canada_data <- rawToChar(canada_raw$content)
canada_tidy <- fromJSON(canada_data)
canada_final <- canada_tidy %>% select(country, year, gdp_per_capita_nominal)
```


**Writing function**

Next, using the methods we used for Canada as a blueprint, I'll create a function to get the data for all countries.
```{r}
#| eval: false
get_gdp_data <- function(country, year, key = my_key) {
  if (is.na(country) || is.na(year)) {
    return(NULL)
  }

  country <- str_replace_all(country, " ", "%20")
  base_url <- "https://api.api-ninjas.com/v1/gdp"
  url <- str_c(base_url, "?country=", country, "&year=", year)
  raw_data <- GET(url, add_headers("X-Api-Key" = my_key))
  json_data <- rawToChar(raw_data$content)
  tidy_data <- fromJSON(json_data)
  if (!is.data.frame(tidy_data) || nrow(tidy_data) == 0) {
    return(NULL)
  }
  if (!"gdp_per_capita_nominal" %in% names(tidy_data) || 
      is.na(tidy_data$gdp_per_capita_nominal)) {
    return(tibble(
      country = country,
      year = year,
      gdp_per_capita_nominal = NA
    ))
  }
  gdps <- tidy_data %>% select(country, year, gdp_per_capita_nominal)
}

```

Let's test the function using USA in 2019.
```{r}
#| eval: false
usa <- get_gdp_data("United States", "2019")
get_gdp_data("hello", "yes")
```

Now we need to apply the function to all countries and years in our measles data.
```{r}
#| eval: false
apply_data <- cases_year %>% select(country, year) %>%
  filter(!is.na(country)) %>% filter(!is.na(year))

# results1 <- map2_dfr(.x = apply_data$country[1:500], .y = apply_data$year[1:500], .f = get_gdp_data)
# results2 <- map2_dfr(.x = apply_data$country[501:1000], .y = apply_data$year[501:1000], .f = get_gdp_data)
# results3 <- map2_dfr(.x = apply_data$country[1001:1500], .y = apply_data$year[1001:1500], .f = get_gdp_data)
# results4 <- map2_dfr(.x = apply_data$country[1501:2000], .y = apply_data$year[1501:2000], .f = get_gdp_data)
# results5 <- map2_dfr(.x = apply_data$country[2001:2380], .y = apply_data$year[2001:2380], .f = get_gdp_data)

results1$country_full <- countrycode(results1$country, origin = "iso3c", destination = "country.name")
results2$country_full <- countrycode(results2$country, origin = "iso3c", destination = "country.name")
results3$country_full <- countrycode(results3$country, origin = "iso3c", destination = "country.name")
results3[98, 4] <- "Pakistan"
results4$country_full <- countrycode(results4$country, origin = "iso3c", destination = "country.name")
results4[101:110, 4] <- "Montenegro"
results5$country_full <- countrycode(results5$country, origin = "iso3c", destination = "country.name")

GDP_APIdata <- bind_rows(results1, results2, results3, results4, results5)
write.csv(GDP_APIdata, "GDPdata.csv", row.names = FALSE)
```

**Brief description of new dataset**

We now have a new dataframe (which I've named "results") which contains 2380 rows with 3 variables. The 2380 rows correlate to the rows in "cases_year" (our original measles dataset) that have both country and year data. 2 of the variables in our new dataframe are country and year, while the last variable is "gdp_per_capita_nominal". As the name suggests, this column contains the nominal GDP per capita for that country, in that year. This was taken from API Ninja's dataset.


## First acessing the API about air quality worldwide

```{r}
#| eval: false
#https://api.api-ninjas.com/v1/airquality?city=New%20York%20City&country=United%20States

my_key = "YOUR_API_KEY_HERE"

base_url <- "https://api.api-ninjas.com/v1/airquality?"

city <- "city=New%20York%20City"
country <- "country=United%20States"

us_url <- glue("{base_url}{city}&{country}")


us_row <- GET(us_url, add_headers("X-Api-Key" = my_key))

us_data <- fromJSON(rawToChar(us_row$content))

```

## Extracting the variable total_aqi to get the total air quality index for each country

```{r}
#| eval: false
us_total <- us_data$overall_aqi

us_total
```

## Creating a dataset that has a city for every country in the measles dataset

```{r}
#| eval: false
cases_year$country[cases_year$year == "2025"]

country_city_tbl <- tibble(
  country = c(
    "Algeria","Benin","Botswana","Burkina Faso","Burundi","Cameroon",
    "Central African Republic","Chad","Comoros","Congo","Côte d'Ivoire","Democratic Republic of the Congo","Equatorial Guinea", "Eswatini","Ethiopia","Gabon","Gambia","Ghana","Guinea","Guinea-Bissau","Kenya","Lesotho","Liberia","Madagascar","Malawi","Mali","Mauritania","Mozambique","Namibia","Niger","Nigeria","Rwanda","Senegal","Sierra Leone","South Africa","South Sudan","Togo","Uganda","United Republic of Tanzania","Zambia","Zimbabwe", "Antigua and Barbuda", "Argentina","Bahamas","Barbados","Belize","Bolivia (Plurinational State of)","Brazil","Canada","Chile","Colombia","Costa Rica","Cuba","Dominican Republic","Ecuador","El Salvador", "Guatemala","Guyana","Haiti","Honduras","Jamaica","Mexico","Nicaragua","Panama","Peru","Saint Lucia","Saint Vincent and the Grenadines", "Trinidad and Tobago","United States of America", "Afghanistan", "Bahrain","Djibouti","Iran (Islamic Republic of)","Iraq","Jordan", "Kuwait","Lebanon","Libya","Morocco","Oman","Pakistan","Qatar","Saudi Arabia","Somalia","Sudan","Syrian Arab Republic","Tunisia","United Arab Emirates","Yemen","Albania","Armenia","Austria","Azerbaijan","Belarus","Belgium","Bosnia and Herzegovina" ,"Bulgaria", "Croatia", "Cyprus", "Czechia","Denmark","Estonia","Finland","France","Georgia","Germany","Greece","Hungary","Ireland","Israel","Italy","Kazakhstan","Kyrgyzstan","Lithuania","Luxembourg","Montenegro","Netherlands (Kingdom of the)","Norway","Poland","Portugal","Republic of Moldova","Romania","Russian Federation", "Serbia", "Slovakia", "Slovenia","Spain","Sweden","Switzerland","Tajikistan","Türkiye","Turkmenistan","Ukraine","United Kingdom of Great Britain and Northern Ireland","Uzbekistan","Bangladesh","Bhutan","Democratic People's Republic of Korea","India","Indonesia","Maldives","Myanmar","Nepal","Sri Lanka","Thailand","Timor-Leste","Australia","Brunei Darussalam", "Cambodia","China","Cook Islands","Fiji","Japan","Kiribati","Lao People's Democratic Republic","Malaysia","Marshall Islands", "Micronesia (Federated States of)","Mongolia","Nauru","New Zealand","Niue","Palau","Philippines","Republic of Korea", "Singapore","Tuvalu","Viet Nam"
  ),
  
  city = c(
    "Algiers","Cotonou","Gaborone","Ouagadougou","Bujumbura","Douala",
    "Bangui","N'Djamena","Moroni","Brazzaville","Abidjan","Kinshasa",
    "Malabo","Mbabane","Addis Ababa","Libreville","Banjul","Accra",
    "Conakry","Bissau","Nairobi","Maseru","Monrovia","Antananarivo",
    "Lilongwe","Bamako","Nouakchott","Maputo","Windhoek","Niamey",
    "Lagos","Kigali","Dakar","Freetown","Johannesburg","Juba","Lomé",
    "Kampala","Dar es Salaam","Lusaka","Harare","St. John's","Buenos Aires","Nassau","Bridgetown","Belize City","Santa Cruz","São Paulo","Toronto","Santiago","Bogotá","San José","Havana","Santo Domingo","Guayaquil","San Salvador","Guatemala City", "Georgetown", "Port-au-Prince","Tegucigalpa","Kingston","Mexico City", "Managua", "Panama City","Lima","Castries","Kingstown","Port of Spain","New York","Kabul","Manama","Djibouti","Tehran","Baghdad","Amman","Kuwait City","Beirut","Tripoli","Casablanca","Muscat","Karachi","Doha","Riyadh","Mogadishu","Khartoum","Damascus","Tunis","Dubai","Sana'a","Tirana","Yerevan","Vienna","Baku","Minsk","Brussels","Sarajevo","Sofia","Zagreb","Nicosia","Prague","Copenhagen","Tallinn","Helsinki","Paris","Tbilisi","Berlin","Athens","Budapest","Dublin","Tel Aviv", "Rome", "Almaty","Bishkek","Vilnius","Luxembourg","Podgorica","Amsterdam","Oslo","Warsaw","Lisbon","Chișinău","Bucharest","Moscow","Belgrade","Bratislava","Ljubljana","Madrid","Stockholm","Zurich","Dushanbe","Istanbul","Ashgabat","Kyiv","London","Tashkent","Dhaka","Thimphu","Pyongyang","Mumbai","Jakarta","Malé","Yangon","Kathmandu","Colombo","Bangkok","Dili","Sydney","Bandar Seri Begawan","Phnom Penh","Shanghai","Avarua",
"Suva","Tokyo","Tarawa","Vientiane","Kuala Lumpur","Majuro","Palikir",
"Ulaanbaatar","Yaren","Auckland","Alofi","Ngerulmud","Manila","Seoul",
"Singapore","Funafuti","Ho Chi Minh City"
  ))
```

## Creating a function to extract total_aqi for each country


```{r}
#| eval: false
get_aqi <- function(city, country){
  
  base_url <- "https://api.api-ninjas.com/v1/airquality?"
  
  city <- paste0("city=", gsub(" ", "%20", city))
  
  country <- paste0("country=", gsub(" ", "%20", country))

  my_url <- glue("{base_url}{city}&{country}")

  my_key = "YOUR_API_KEY_HERE"
  
  my_row <- GET(my_url, add_headers("X-Api-Key" = my_key))

  my_data <- fromJSON(rawToChar(my_row$content))
  
  if (length(my_data$overall_aqi) == 0){
    
    my_total_aqi <- NA
    
  }else {
    
    my_total_aqi <- my_data$overall_aqi
    
  }
  
  return(my_total_aqi)
  
}
```

## Testing the function on Algiers, Algeria
```{r}
#| eval: false
get_aqi("Algiers", "Algeria")
```

## Running the function over the full dataset and creating a new data frame

```{r}
#| eval: false
first_run <- map2(.x = country_city_tbl$city, .y = country_city_tbl$country, .f = ~get_aqi(city = .x, country = .y))

aqi_df <- data.frame(total_aqi = unlist(first_run))

nrow(country_city_tbl)
nrow(aqi_df)

new <- cbind(country_city_tbl, aqi_df)
new

write.csv(new, "new.csv")
```

## Data description: Air Quality Index

This is a data frame that contains every total Air Quality Index by city and
country in the year 2025. There are 168 rows for the 168 countries in the measles
data but two cities have no data: Niamey, Niger and Seoul, Republic of Korea. 

#### Read in CSV

```{r}
url <- "https://raw.githubusercontent.com/amandaadong/project1/main/GDPdata.csv"
gdp_df <- read.csv(url)
head(gdp_df)
results <- read_csv("https://raw.githubusercontent.com/amandaadong/project1/main/GDPdata.csv")
new <- new <- read_csv("https://raw.githubusercontent.com/nataliecc686/measles-analysis/main/new.csv")
```

#### Join all the datasets (meta)

```{r}
results_clean <- results |>
  mutate(
    country = str_replace_all(country, "%20", " "),
    year = as.numeric(year)
  )

meta_df <- cases_year |>
 left_join(new, by = "country") |>
 left_join(gdp_df, by = c("iso3" = "country", "year" = "year"))
```

### Research Question: How does the relationship between GDP per capita and measles incidence rate change before and after COVID?

```{r}
meta_df |>
  mutate(
    time_period = case_when(
      year >= 2015 & year <= 2019 ~ "Pre-COVID (2015-2019)",
      year >= 2020 & year <= 2023 ~ "Post-COVID (2020-2023)"),
      time_period = factor(time_period, levels = c("Pre-COVID (2015-2019)", "Post-COVID (2020-2023)"))) |>
  filter(!is.na(time_period)) |>
  group_by(iso3, country_full, region, time_period) |>
  summarize(
    avg_incidence = mean(measles_incidence_rate_per_1000000_total_population, na.rm = TRUE),
    avg_gdp = mean(gdp_per_capita_nominal, na.rm = TRUE)) |>
  filter(!is.na(avg_incidence), !is.na(avg_gdp), 
         avg_incidence < 500, avg_gdp > 0) |> #removed extreme outliers to observe differences
  ggplot(aes(x = avg_gdp, y = avg_incidence, color = region)) +
  geom_point(alpha = 0.5, size = 2) +
  facet_wrap(~time_period) +
  labs(
    title = "Pre-COVID and Post-COVID Periods vs GDP",
    x = "Average GDP per Capita",
    y = "Average Measles \nIncidence per Million",
    color = "WHO Region") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0),
        plot.title.position = "plot",
        axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title.y = element_text(angle = 0, vjust = 0.5),
        panel.grid.minor = element_blank(),
        legend.position = "top",
        legend.title = element_text(face = "bold")
        )
```

### Visualization: Average AQI vs. Average Measles Cases by Country (2020–2025)

```{r}
#| message: false
#| warning: false
meta_df %>%
  filter(!is.na(total_aqi), !is.na(total_suspected_measles_rubella_cases)) %>%
  group_by(country, total_aqi) %>%
  summarise(avg_cases = mean(total_suspected_measles_rubella_cases, na.rm = TRUE)) %>%
  ggplot(aes(x = total_aqi, y = avg_cases)) +
  geom_point(alpha = 0.6, color = "dodgerblue") +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  scale_y_log10(labels = scales::comma) +
  labs(
    title = "Air Quality Index vs. Average Measles Cases by Country",
    subtitle = "Each point represents one country; cases shown on log scale",
    x = "Air Quality Index (higher = worse air quality)",
    y = "Average Annual Measles Cases (log scale)",
    caption = "Sources: API Ninjas AQI data; TidyTuesday measles dataset"
  )
```
```{r}
aqi_model <- meta_df |>
  filter(!is.na(total_aqi), !is.na(total_suspected_measles_rubella_cases)) |>
  group_by(country, total_aqi) |>
  summarise(avg_cases = mean(total_suspected_measles_rubella_cases, na.rm = TRUE),
            .groups = "drop") |>
  mutate(log_avg_cases = log10(avg_cases + 1))

# Fitting the model
model <- lm(log_avg_cases ~ total_aqi, data = aqi_model)
summary(model)
```


### Research Question: How does the relationship between GDP per capita and measles incidence rate change before and after COVID?

```{r}
meta_df |>
  mutate(
    time_period = case_when(
      year >= 2016 & year <= 2019 ~ "Pre-COVID (2016-2019)",
      year >= 2020 & year <= 2023 ~ "Post-COVID (2020-2023)"),
      time_period = factor(time_period, levels = c("Pre-COVID (2016-2019)", 
                                                   "Post-COVID (2020-2023)"))) |>
  filter(!is.na(time_period)) |>
  group_by(iso3, country_full, region, time_period) |>
  summarize(
    avg_incidence = mean(measles_incidence_rate_per_1000000_total_population, na.rm = TRUE),
    avg_gdp = mean(gdp_per_capita_nominal, na.rm = TRUE)) |>
  filter(!is.na(avg_incidence), !is.na(avg_gdp), 
         avg_incidence < 500, avg_gdp > 0) |> #removed extreme outliers to observe differences
  ggplot(aes(x = avg_gdp, y = avg_incidence, color = region)) +
  geom_point(alpha = 0.5, size = 2) +
  facet_wrap(~time_period) +
  labs(
    title = "Pre-COVID and Post-COVID Periods vs GDP",
    x = "Average GDP per Capita",
    y = "Average Measles \nIncidence per Million",
    color = "WHO Region") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0),
        plot.title.position = "plot",
        axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title.y = element_text(angle = 0, vjust = 0.5),
        panel.grid.minor = element_blank(),
        legend.position = "top",
        legend.title = element_text(face = "bold")
        )
```



