---
title: "Measles Global Analysis"
format:
  dashboard:
    orientation: columns
    embed-resources: true
---

```{r}
#| label: packages-setup-and-load-data
#| message: false
#| warning: false
library(httr)
library(tidyjson)
library(jsonlite)
library(tidyverse)
library(stringr)
library(gt)
library(paletteer)
library(ggridges)
library(broom)

cases_month <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-06-24/cases_month.csv')
cases_year  <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-06-24/cases_year.csv')
meta_df     <- read_csv("https://raw.githubusercontent.com/amandaadong/project1/main/metadf.csv")
```

```{r}
#| label: compute-value-box-values
#| message: false
#| warning: false

# --- RQ1 value box inputs ---
# TODO: Replace column names with actual names from cases_year
# yearly_totals  <- cases_year |>
#   group_by(year) |>
#   summarise(total = sum(cases, na.rm = TRUE))
#
# highest_year   <- yearly_totals |> slice_max(total, n = 1)
# lowest_year    <- yearly_totals |> slice_min(total, n = 1)
# max_pct_change <- yearly_totals |>
#   mutate(pct_change = (total - lag(total)) / lag(total) * 100) |>
#   slice_max(abs(pct_change), n = 1) |>
#   pull(pct_change)

# --- RQ2 regression inputs ---
# TODO: Replace aqi / country column names with actual names from meta_df
# rq2_df <- meta_df |>
#   left_join(
#     cases_year |>
#       group_by(country) |>
#       summarise(avg_cases = mean(cases, na.rm = TRUE)),
#     by = "country"
#   ) |>
#   mutate(log_cases = log10(avg_cases)) |>
#   filter(!is.na(aqi), !is.na(log_cases))
#
# lm_fit  <- lm(log_cases ~ aqi, data = rq2_df)
# lm_sum  <- summary(lm_fit)
# lm_coef <- tidy(lm_fit, conf.int = TRUE)
#
# p_value <- lm_sum$coefficients["aqi", "Pr(>|t|)"]
# t_value <- lm_sum$coefficients["aqi", "t value"]
# ci_low  <- lm_coef |> filter(term == "aqi") |> pull(conf.low)
# ci_high <- lm_coef |> filter(term == "aqi") |> pull(conf.high)
#
# --- RQ2 AQI subsets for tables ---
# top10_aqi    <- rq2_df |> slice_max(aqi, n = 10)
# bottom10_aqi <- rq2_df |> slice_min(aqi, n = 10)
#
# --- RQ2 AQI quartiles for Figure 4 ---
# rq2_df <- rq2_df |>
#   mutate(aqi_quartile = ntile(aqi, 4) |>
#            factor(labels = c("Q1 (Low AQI)", "Q2", "Q3", "Q4 (High AQI)")))
```

# RQ1: Global Trends Over Time {orientation="columns"}

## Column - Table {width="45%"}

### Row - Table 1

::: {.card title="Table 1: Global Measles Cases by Year"}
```{r}
#| label: gt-table-rq1
# --code-- Insert table 1 of global cases by year
```
:::

### Row - Table 2

::: {.card title="Table 2: Detection Method Proportions"}
```{r}
#| label: gt-table-detection
# --code-- Insert table 2 of detection method prop
```
:::

## Column - Value Boxes & Chart {width="55%"}

### Row - RQ1 Value Boxes {height="25%"}

::: {.valuebox icon="arrow-up-circle" color="danger"}
Highest single-year cases

`r "TODO: paste(highest_year$year, '—', format(highest_year$total, big.mark=','))"`
:::

::: {.valuebox icon="arrow-down-circle" color="success"}
Lowest single-year cases

`r "TODO: paste(lowest_year$year, '—', format(lowest_year$total, big.mark=','))"`
:::

::: {.valuebox icon="percent" color="warning"}
Largest year-over-year % change

`r "TODO: paste0(round(max_pct_change, 1), '%')"`
:::

### Row - Figure 1 {height="37%"}

::: {.card title="Figure 1: Global Measles Cases Over Time"}
```{r}
#| label: rq1-trend-chart
# --code-- lollipop figure 1: Global Measles Cases Over Time
```
:::

### Row - Figure 2 {height="38%"}

::: {.card title="Figure 2: Detection Method by Year"}
```{r}
#| label: rq1-detection-chart
# --code-- stacked bar charts Figure 2: Detection Method by Year
```
:::

# RQ2: Air Quality & Measles Cases {orientation="columns"}

## Column - Figures {width="50%"}

### Row - Figure 3 {height="50%"}

::: {.card title="Figure 3: Linear Regression — AQI vs. log10 Average Measles Cases"}
```{r}
#| label: rq2-lm-scatter
# --code-- Figure 3: Linear Regression between Air Quality Index and log10 transformed Average Measles Cases by Country
```
:::

### Row - Figure 4 {height="50%"}

::: {.card title="Figure 4: Measles Distribution by AQI Quartile"}
```{r}
#| label: rq2-aqi-quartile
# --code-- Figure 4: Measles Distribution by AQI Quartile
```
:::

## Column - Table, Value Boxes & Subset Tables {width="50%"}

### Row - Table 5 {height="20%"}

::: {.card title="Table 5: Linear Regression Coefficients"}
```{r}
#| label: rq2-lm-table
# --code-- Table 4: Linear Regression Coefficients
```
:::

### Row - RQ2 Value Boxes {height="15%"}

::: {.valuebox icon="calculator" color="info"}
p-value (AQI slope)

`r "TODO: format(p_value, digits = 4, scientific = TRUE)"`
:::

::: {.valuebox icon="bar-chart" color="warning"}
t-value (AQI slope)

`r "TODO: round(t_value, 3)"`
:::

::: {.valuebox icon="arrows-expand" color="secondary"}
95% Confidence Interval

`r "TODO: paste0('[', round(ci_low, 3), ', ', round(ci_high, 3), ']')"`
:::

### Row - AQI Subset Tables {height="65%"}

::: {.panel-tabset}

## Top 10 — Highest Avg AQI
```{r}
#| label: rq2-top10-aqi
#--code-- gt table of top 10 countries by average AQI using Table 3: AQI vs. Average Measles Cases (unhealthy air quality rating)

```

## Bottom 10 — Lowest Avg AQI
```{r}
#| label: rq2-bottom10-aqi
#--code-- gt table of bottom 10 countries by average AQI using Table 3: AQI vs. Average Measles Cases (best air quality rating)
```

:::
